---
title: "Course Project 1"
author: "Kayla Maxwell"
date: "May 9, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

## Loading the Data

Code to download and read the data into R
```{r data}
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
file <- unzip(temp)
unlink(temp)
  
data <- read.csv(file, header = TRUE, sep = ',', colClasses = c("numeric", "character",
                                                                          "integer"))
```
  
## Process the Data
Change date format using lubridate

```{r}
library(lubridate)
data$date <- ymd(data$date)
```

## Analyze the Data

**What is the mean total number of steps taken per day?**

1.Calculate the total number of steps taken per day

```{r}
totalsteps1 <- tapply(data$steps,data$date,sum,na.rm=TRUE)
```

2. Make a histogram of the total number of steps taken each day
```{r}
hist(totalsteps1, xlab="Steps", main="Total Steps Per Day" )
```


3.Calculate and report the mean and median of the total number of steps taken per day
```{r}
mean(totalsteps1)
median(totalsteps1)
```
**What is the average daily activity pattern?**
1.Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}


##Find average steps per interval

avgperinterval <- tapply(data$steps,data$interval,mean,na.rm=TRUE)

##Create line plot
plot(row.names(avgperinterval), avgperinterval, type="l", xlab="Interval", ylab="Average Number of Steps", main="Average Number of Steps per Interval")
```
2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

Interval **835** at row 104

```{r}
which.max(avgperinterval)

```
**Imputing Missing Values**
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).

```{R}
nrow(data[is.na(data),])
```

2.Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.



```{R}
## na data
nadata <- data[is.na(data),]

## complete data
cleandata <- data[complete.cases(data),]
## replace nas with average per interval
nadata$steps <- as.numeric(avgperinterval)

```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
##combine datasets
newdata <- rbind(nadata, cleandata)
newdata <- newdata[order(newdata[,2],newdata[,3]),]
```

4.Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

The values are different. Imputing missing data shifted the average and median of the data.

```{r}
##new total steps by day
newtotal <- tapply(newdata$steps,newdata$date,sum)
hist(
        newtotal,
        col = "purple",
        main = "Adjusted Histogram of the Total Number of Steps",
        xlab = "Total Number of Steps",
        breaks = 5
)

##calculate mean and median
mean(newtotal)
median(newtotal)
```
**Are there differences in activity patterns between weekdays and weekends?**

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
##create new variable for weektype
library(dplyr)
newdata<- newdata%>%
        mutate(weektype= ifelse(weekdays(newdata$date)=="Saturday" | weekdays(newdata$date)=="Sunday", "Weekend", "Weekday"))
```
2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r}
library(ggplot2)
Interval<- newdata%>%
        group_by(interval, weektype)%>%
        summarise(avg= mean(steps, na.rm=TRUE))

plot<- ggplot(Interval, aes(x =interval , y=avg, color=weektype)) +
       geom_line() +
       labs(title = "Average Daily Steps by Weektype", x = "Interval", y = "Number of Steps") +
       facet_wrap(~weektype, ncol = 1, nrow=2)
print(plot)
```
Daily steps are higher in the earlier part of the day during the week, but higher overall on the weekends. 
